# Future value

The Powerball is a weekly lottery famous for its outsized payoff. For instance, for the week of April 7, 2021, the jackpot payout is officially described as $43,000,000. There is a limited sense in which this might be true, but in a far more meaningful sense it is not. The winner has a choice between a one-time cash payout of $29,300,000 or an annuity of 30 annual payments of $1,433,333. However, the winnings are taxable income and the lottery's administration (a not-for-profit, government benefit organization) withhold applicable federal and state taxes. According to the [usmega site](https://www.usamega.com/powerball/jackpot), in Colorado the cash payment would amount to $17,161,778 while the annuity would be 30 payments of $873,711. `r mark(4300)`

This will sound silly, but we know the value of $17,161,778 paid today: it is $17,161,778. But what is the value of $873,711 paid out over 30 years? The answer depends on the *discount rate*, that is, the present value of future money. When the discount rate is $r$, the current value of money due to be paid in $n$ years is $(1-r)^n$. For instance, at $r = 5\%$, the current value of $1000 to be paid 10 years in the future is $599. The table shows the net present value of the $873,711 paid out yearly over 30 years. `r mark(4305)`

Discount rate | Net Present Value
--------------|------------------:
0    | `r format(sum(1^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
1%   | `r format(sum(.99^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
2%   | `r format(sum(.98^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
4%   | `r format(sum(.96^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
5%   | `r format(sum(.95^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
8%   | `r format(sum(.92^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
10%  | `r format(sum(.90^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`
15%  | `r format(sum(.85^(0:29))*873711, digits=4, scientific=FALSE, big.mark=",")`

The formula behind this table is $$873711 \sum_{k=0}^{29} (1-r)^k$$

What discount rate to use is a judgement call, and will reasonably depend on a person's situation. For example, a 90-year old might reasonably discount future money at $r=20\%$, whereas the discount rate for a person owing money next month to the mafia might be considerably higher.  `r mark(4310)`

As a placeholder, we can look to the auctions of inflation-adjusted US Treasury Inflation Protected Securities (TIPS). As of this writing, a TIPS with a nominal payout of $1000 in 30 years is for sale at $100.70 (as of April 6, 2021), corresponding to a discount rate of 8% per year. With some justification, one might assert that the nominal $43,000,000 Powerball prize is really worth about $10 million. `r mark(4315)`

Let's turn the problem of calculating a net present value into a calculus problem. Rather than summing over the discounted yearly dollar payout, we'll consider a continuous payout rate in dollars per year.  `r mark(4320)`

```{r pb1-1, echo=FALSE, results="markup"}
askMC(
  "In continuous time, a continuously compounding interest rate $r$ is a parameter in a differential equation $$\\dot{P} = r P$$ where $P$ is the amount of money (\"principal\") under investment.  Solve the differential equation at $t=1$ year  with an initial principal of $100  to find the corresponding *effective annual interest rate*. Which of these is the conversion between $r$ in the differential equation and the effective annual interest rate? (Hint: Think carefully about the distinction between the interest rate $r$ and the increase at the end of one year $1+r$)",
  "+$\\exp(r) - 1$+",
  "$\\exp(1+r) -1$",
  "$\\exp(r)$",
  "$\\exp(1+r)$"
)
```

```{r pb1-2, echo=FALSE, results="markup"}
askMC(
  "Using the correct answer from the previous question, find the continuously compounding interest rate $r$ that corresponds to an effective annual discount rate of 8%. Which of these is it?",
  "$r = \\ln(0.08)$",
  "+$r = \\ln(1.08)$+",
  "$r = \\sqrt{0.08}$",
  "$r = \\sqrt{(1 + 0.08)^2} - 1$"
  
)
```


```{r pb1-3, echo=FALSE, results="markup"}
askMC(
  "Which of these integrals corresponds to the discrete sum in the formula for the net present value of the Powerball payout?",
  "+$\\int_0^{29} 873711 (1-r)^t dt$+",
  "$\\int_1^{30} 873711 (1-r)^t dr$",
  "$\\int_0^{30} 873711 (1-r)^t dt$",
  "$\\int_1^{29} 873711 (1-r)^t dr$"
)
```


Here is a sandbox you can use for the calculations needed to answer the following questions.

```{r pb2-1, exercise=TRUE, exercise.cap="Calculation sandbox", exercise.nlines=10, eval = FALSE}




```


```{r answer-to-last-question, echo=FALSE, eval=FALSE}
F <- antiD((1-r)^t ~ t, r=0.077)
10*(F(100000) - F(0))
```


We're going to use an effective annual discount rate of 8% in the following problems.

```{r pb1-4, echo=FALSE, results="markup"}
askMC(
  "Which of the following is the continuously compounded interest rate that corresponds to the effective annual interest rate of 8%?",
  "7.3%",
  "+7.7%+",
  "8.0%", 
  "8.3%",
  random_answer_order = FALSE
)
```

```{r pb1-5, echo=FALSE, results="markup"}
askMC(
  "Using the continously compounded interest rate that's equivalent to an effective annual rate of 8%, calculate the appropriate integral for the net present value of a continuous payout at a rate of $873,711 dollars/year$ for the 30-year annuity term used in Powerball. Which of these is it? (Hint: Construct the appropriate anti-derivative and from that compute the appropriate definite integral.)", 
  "$8,970,000",
  "+$9,840,000+",
  "$10,030,000",
  "$10,460,000",
  random_answer_order = FALSE
)
```

There is a joke that makes sense only to the financially savvy: When the Powerball claims a $1 million payout, they mean $1 per year over a million years. We can do this calculation using numerical evaluation of the definite integral but there is a catastrophic loss of numerical precision for the 1-million year calculation. Instead, do the calculation for a $10/year payout over 100,000 years. `r mark(4325)`

First, answer the following **ungraded** question with utter honesty.

```{r pb1-6a, echo=FALSE, results="markup"}
etude2::etudeEssay(
  "**Before doing the actual calculation**, form an intuitive answer to the question of what is the net present value of $10/year over 100,000 years. Don't try to be strategic in anticipating the answer, just give your gut feeling. The instructors are curious about what sorts of numbers will come to mind." `r mark(4330)`
)
```

```{r pb1-6b, echo=FALSE, results="markup"}
askMC(
  "At the continuously compounded interest rate corresponding to an effective annual rate of 8%, find the net present value of $10 per year over a hundred-thousand years. Which one of these is closest?",
  "+$124.80+",
  "$1248.00$",
  "$12,480.00$",
  "$124,800.00$",
  random_answer_order = TRUE
)
```

Finally, and this has nothing to do with net present value *per se*, try out the calculation of the net present value of $1 per year over a million years. This is worth doing if only to see a compelling example of how far off computer numerics can be when using very large or very small numbers. `r mark(4335)`

A professional will always have at hand some method for checking the reasonableness of an answer in order to detect when something is wrong. For example, here you might look at a sequence of calculations: $100,000 per year over 10 years, $10,000 per year over 100 years, and so on. You would expect the numbers in this sequence to approach some sort of limit as the number of years increases. It does, but as the number of years gets larger that limit goes awry. `r mark(4340)`


