# Projection & decomposition

In the previous chapter, we took linear combinations of vectors. Using the notation of ***matrices***, we solved the problem
$$\text{previous chapter:}\ \ {\mathbf A}\, \vec{x} = \vec{?}$$ where $\vec{x}$ is a column of coefficients, one coefficient for each of the columns in the matrix ${\mathbf A}$. Finding $\vec{?}$ is a simple, if somewhat tedious, matter of multiplication. 

$$\text{this chapter:}\ \ {\mathbf A}\, \vec{?} = \vec{b}$$
In other words, the task of this chapter is to show how to solve ${\mathbf A} \vec{x} = \vec{b}$ when we know the result of the linear combination $\vec{b}$, and the set of vectors being combined, $\mathbf M$, but we don't yet know $\vec{x}$. 

As motivation for this "find $\vec{x}$" problem, we refer you to Figure \@ref(fig:water-dots) which showed the temperature-vs-time data from the `CoolingWater` data frame. That figure shows several possible linear combinations of the vectors $u(t) \equiv e^{-0.02 t}$ (we called this `expkt`) and $v(t) \equiv 1$ (we called this `intercept`. Suppose we seek to find the particular linear combination of $u(t)$ and $v(t)$ that comes as close as possible to the black dots in the figure. That is, we know ${\mathbf A}$: the two columns `expkt` and `intercept` from the data frame, and we know $\vec{b}$: the column `temp` from the data frame. 

This sort of problem is extremely common and important throughout quantitative fields of all sorts, from astronomy to zoology, and is one of the foundation techniques in statistics and data science. We'll present the approach both graphically and algorithmically. 

## The residual vector



## Finding the target with QR 

Start with the very simple example of the identity matrix and reaching a target with that.

Show that the vectors in the identity matrix are mutually orthogonal.

Show that we can find a mutually orthogonal basis set for any matrix.


## Categorical inputs

Why a categorical variable corresponds to a set of indicator vectors. Why it wouldn't work to assign numbers 1,2,3,... to the categories (because that would presume that the second category is mathematical between the first and third).

Refer back to the data frame `mtcars` [BUT THIS IS ALL NUMERIC, SO NOT A GOOD SOURCE FOR CATEGORICAL EXAMPLES] displayed on a previous page. Describe how `model.matrix()` used the data to create the right-most vector in the model matrix.

## Exercises

`r insert_calcZ_exercise("XX.XX", "wniqMc", "Exercises/panda-begin-shoe.Rmd")`

`r insert_calcZ_exercise("XX.XX", "yeY17y", "Exercises/kangaroo-light-bowl.Rmd")`


`r insert_calcZ_exercise("XX.XX", "p209w8", "Exercises/beech-build-shirt.Rmd")`


`r insert_calcZ_exercise("XX.XX", "dvWdE7", "Exercises/puppy-dream-spoon.Rmd")`


a. Calculate the angle between two vectors
    i. arithmetically, using dot product
    ii. graphically, using protractor
b. Find a nonzero vector orthogonal to a given vector
c. Calculate the projection of a vector onto another vector
    i. arithmetically with dot product
    ii. graphically
d. Decompose a vector into a residual and a component directed along a second vector
    i. graphically
    ii. arithmetically



Our goal is to scale the `expkt` vector so that the scaled numbers will be as close as possible to our destination, namely, `temp`. Comparing the two columns of numbers, you might anticipate that the scalar will be about 100. We'll see how to calculate it exactly in the next chapter. The result turns out to be 99.23. The resulting model will be $$T(t) = 99.23\, e^{-0.02 t}\ .$$

How are we to judge whether this is a good model or not? Common sense suggests plotting out the model function along with the data, as in Figure \@ref(fig:just-exp).

```{r just-exp, echo=FALSE, fig.cap="Comparing the model $99.23\\, e^{-0.02 t}$ to the recorded data in `CoolingWater."}
gf_point(temp ~ time, data = CW) %>%
  slice_plot(99.23*exp(-0.02*time) ~ time, color="magenta")

```

Judge for yourself whether this is a good model. The obvious deficiency is that the model falls, as decaying exponentials will do, toward a temperature of 0, whereas the water is cooling to a room temperature of about 25 degrees.

Let's return to the model seen in terms of vectors. The advantage of doing this is to develop a general procedure we can use for interpreting models of all sorts, rather than just the particular situation of the cooling-water data.

What are the geometric facts? We know that the `temp` vector has length 251.3 deg C. Similarly we can calculate the length of the `expkt` vector: 2.46 deg C. 

It might seem that the "direction" of the vector is meaningless, because it's a direction in an abstract, hard-to-envision 15-dimensional space. (There are 15 components to each of `temp` and `expkt`.) Even so, we can calculate the ***angle*** between the two vectors, using the formula $\cos(\theta) = \frac{\vec{v}\cdot \vec{w}}{\|\vec{v}\|\ \|\vec{w}\|}$. Doing the arithmetic gives $$\cos{\theta} = \frac{599.8}{251.3 \times 2.46} = 0.9708\ \ \implies \ \ \ \theta = 13.88^\circ$$

```{r temp-expkt-picture, echo=FALSE, fig.cap="The vectors `temp` and `expkt` have an angle of 13.88 deg between them. Here, `expkt` has been drawn 10x it's actual size."}
gvec(from = c(0,0), to = c(250, 25.1), color = "black") %>%
  gf_segment(0 + 3*28.8 ~ 0 + 3*116.5, 
             color="magenta", linetype = "dotted") %>%
  gvec(from=c(0,0), to = 2*c(11.6, 2.88), color="magenta") %>%
  gf_refine(coord_fixed())
```

With these geometrical facts, we can draw a picture. Figure \@ref(fig:temp-expkt-picture) shows `temp` in black and `expkt` in magenta. (We've drawn it 10 times as long as it really is so that you can see it well.) For the `expkt` vector to be a good model of `temp`, we need to scale it so that the result, which must be on the dotted line in the picture, is as close as possible to the tip of `temp`. You can count off yourself how many `expkt` steps will bring you close to `temp`. (Remember to multiply your result by 10, since in the picture we drew `expkt` ten times longer than its arithmetic length.)

One reasonable way to quantify how good a model of `temp` can be made by a properly scaled version of vector `expkt` is the angle between them: 13.88 degrees. 

Likewise, we can scale the vector `intercept` to make it match `temp` as well as possible. The angle between `intercept` and `temp` works out to be 75.7 degrees; the vectors are not very well aligned. Scaling by 58.2 will bring `intercept` as close as it is ever going to get to `temp`, which is not very close at all.

The idea of a linear combination is to scale and add multiple vectors. As a **very rough start**, let's look at the combination 58.2 `intercept` + 99.23 `expkt`, the combination of the two individual models we constructed by vector analogy. CAUTION: The model will be poor. That's not because the vector analogy is poor but because we still have to work out, as we will in the next two chapters, how properly to work with vectors. 

```{r}
CW <- CW %>% 
  mutate(model = 99.23*expkt + 58.2*intercept)
```

```{r echo=FALSE}
knitr::kable(CW) %>% kableExtra::kable_minimal()
```

The resulting model is ... well, terrible! Figure \@ref(fig:two-temp-models) shows the linear combination 

```{r two-temp-models, echo=FALSE}
gf_point(temp ~ time, data = CW) %>%
  gf_point(model ~ time, data = CW, color="magenta") %>%
  slice_plot(58.2+ 99.23*exp(-0.02*time) ~ time, color="magenta")
```

EXERCISE: Repeat the calculations for the entire `CoolingData` data frame.


Adding vectors. The result is a vector

Scaling and adding vectors: a linear combination of vectors.



## MOVED FROM VECTORS




A pencil is a physical object that does a good job representing a vector in three dimensional space. Three-dimensional space is all around us and it's easy---indeed, inevitable---to situate a pencil in it. We have no such physical access to 4-dimensional space or higher-dimensional spaces. Instead of a physical representation, we need to rely on a mathematical one: a column of numbers. And we can't use a protractor to measure the angle between two vectors in 4- or higher-dimensional space. Instead, we calculate the angle using arithmetic. But to define the numerical process for calculating an angle, we need to make sure that the result follows the familiar conventions for angles, specifically that they be between 0 and 180 degrees, that the angle between two vectors with the same orientation is 0, and that the angle between two opposite-pointing vectors is 180 degrees.

The arithmetic formula for computing the angle between two vectors is simple. Or, rather, it's simple if we allow ourselves to calculate the cosine of the angle rather than the angle $\theta$ itself. The underlying quantity can be calculated using dot products: 
$$\cos(\theta) \equiv \frac{\vec{v}^T \cdot \vec{w}}{\sqrt{\strut (\vec{v}^T\cdot \vec{v})(\vec{w}^T\cdot \vec{w})}}$$




A starting  point  is creating a vector on the computer. In  some sense this a vector is just a collection  of numbers, but it's helpful  to  be disciplined  and  remember that, for our purposes, a  vector is  a **column** of numbers. R knows about such columns and will  handle them appropriately.

One way to create a column vector is with  the  `rbind()` function.

use the  `rbind()` function applied to individual arguments. Here,  for instance,  is a command that makes a three-dimensional vector we are calling `b`.

```{r}
b <- rbind(4, -2, 6)
b
```

Notice that in  printing out a  vector, R  includes a series of indices  (e.g. `[1,]` or `[3,]`) to help the reader identify the location  of any element in  the vector. It also  prints a header (`[,1]`) which  is helpful later when we work with  collections of vectors.

## Matrix

You are going to hear the word "matrix" a lot.  Later in  this tutorial we will use the term "matrix multiplication." A matrix is a collection  of  vectors, all of the same dimension. We'll get to them  in good time.  

## Scalar multiplication

You  can multiply a vector times  a  number. The result is a new vector with exactly the  same direction as  the original, but with  a different length. The  arithmetic  is  very  simple: do ordinary multiplication  of the number by  each of the elements  of the  vector. Examples:

$$2 \left(\begin{array}{c}4\\7\end{array}\right) = \left(\begin{array}{c}8\\14\end{array}\right)$$

This  simple  multiplication is called "scalar multiplication" for two  reasons:

1. The result is to "scale" the vector, in the sense of a  "scale  model",  that  is, to make the vector bigger or smaller.
2. There is another important form of vector  arithmetic called "matrix multiplication." By  saying  "scalar multiplication,"  we avoid the  confusion  that might arise if  we  used "multiplication" alone.

In R, scalar multiplication of a vector is  done with `*`, just like ordinary multiplication   with numbers:

```{r echo=TRUE}
b <- rbind(4, -2, 6)
2.3 * b
```

::: {.example data-latex=""}
In a `r sandbox_link()`, Write  the  R  code to create a vector named `w` with  components 4, -1,  and -3.5.  Then scalar multiply `w` by 6.3.

```{r echo=TRUE, eval=FALSE} 
w <- _______
6.3 _______ w
```

<details>
<summary>Solution</summary>
```{r}
w <- rbind(4,  -1,  -3.5)
6.3 *  w
```
</details>
:::

## Dot product and `%*%`

Now to  introduce  a new R arithmetic function, written `%*%`. This symbol is pronounced "matrix multiply."  In traditional mathematical notation, matrix multiplication is indicated by putting the two quantities next to one another, like this: $\vec{\mathbf m}^T \vec{\mathbf x}$, or sometimes with a dot $\vec{\mathbf m}T\cdot \vec{\mathbf x}$. The superscript $^T$ means "transpose." For us, this  is merely a  book-keeping convention.

The operation `%*%` will do several  different types  of  arithmetic  with  vectors. The one we will work with here is  called  a *dot product*.  (There are  also "matrix products"  and "outer products".)

The R notation for a dot product very  much echoes the  traditional matrix  notation,  at least  with  respect to $^T$.  We'll  illustrate by creating  two vectors `u`  and `v` and then calculating their dot  product.

```{r}
u <- rbind(6, -3,  7)
v <- rbind(2,  1,  3)
t(u) %*%  v
```

Notice  that the  output  of a dot product is  a single number: a scalar. (R prints the output as if the scalar were  a vector in one-dimension.)

Arithmetically, the dot product is calculated by multiplying the corresponding components in the two vectors  (e.g.  $6  \times 2$ and  $-3 \times 1$ and  $7 \times 3$) and adding up the  result. You can see why the dot product always involves two vectors with the same number of elements.

The R `t()` function corresponds to  the mathematical notation for the transpose: $^T$. So `t(u)` would be written, mathematically, as $\vec{u}^T$. The  purpose of `t()` is to turn columns (like our vector `u`) into rows,  and vice versa.  If you  like,  try the command `t(u)` in  the sandbox to see how it  is printed.

For us,  the purpose of writing `t(u)` is  to signal  to  the `%*%`  matrix multiplication  operation that we want a particular operation: the dot product.

The  dot product always involves the transpose of a column vector on  the left side of  `%*%`  and a column vector on  the right side.

You  can  also  write a command  `u %*% t(v)`,  but  this is not a  dot product. It is  called  an "outer product" and we will not need it in this course. Try it out in  the sandbox.

::: {.caution}
The ***dot product*** and the ***outer product*** are written in very similar ways but produce completely different results. We won't have much use for outer products in this course, but you should be aware what they look like so that you can diagnose the problem if your attempt at a dot product goes wrong.

**Dot products**
```{r}
t(u) %*% v
t(v) %*% u
```

**Outer product**
```{r}
u %*% t(v)
v %*% t(u)
```
:::


## Vector lengths

For  a vector $\vec{v}$, the  length is denoted $|| \vec{v} ||$. Vector length can  be measured with  a ruler ...  so long as you have physical access to  the vector. But often, all we have  is the numerical representation. So, we use arithmetic---the dot product---to calculate vector length: 
$$|| \vec{v}  || \equiv \sqrt{\ \vec{v}^T \cdot \vec{v}}$$

::: {.example data-latex=""}
Consider the two vectors 
$$\vec{u} \equiv \left(\begin{array}{c}3\\4\end{array}\right) \  \  \ \mbox{and}  \ \ \ \vec{w} \equiv \left(\begin{array}{c}1\\1\\1\\1\end{array}\right)
$$

The length of $\vec{u}$ is $|| \vec{u} || = \sqrt{\strut 3^2 + 4^2} = \sqrt{\strut 25} = 5$.

The length of $\vec{w}$ is $|| \vec{w} || = \sqrt{\strut 1^2 + 1^2 + 1^2 + 1^2} = \sqrt{\strut 4} = 2$.

Using a `r sandbox_link()`, use R commands to create the vectors $\vec{u}$ and $\vec{w}$ and find their lengths using the `dot-product operator`%*%` operator.

```{r eval=FALSE}
u <- rbind( ____ )
# length of u
sqrt( ____ %*% ____ )

v <- rbind( ______ )
# length of v
sqrt( ____ %*% ____ )
```

<details>
<summary>Solution</summary>
```{r}
u <- rbind(3, 4)
sqrt(t(u) %*% u)
w <- rbind(1, 1, 1, 1)
sqrt(t(w) %*% w)
```
</details>



## Outline

a. Write a system of linear equations as a "target problem" involving a linear combination of vectors. 
b. Definition of a linear combination of vectors.
c. Matrix is a collection of vectors
    i. Linear combination is written as matrix multiplied by vector
d. Solve target problem algebraically or graphically in two dimensions.
e. Subtracting vectors algebraically and graphically




