# Vectors

```{r include=FALSE}
library(Znotes)
```

a. Dual representations of vectors.
    i. arithmetic: a vector is a column of numbers.
    ii. graphically: an arrow.
        a.  direction and a magnitude.
        b. Position isn't a feature. You can draw the vector whereever is convenient so long as you keep the direction and magnitude.
b. A vector is embedded ("lives in") a space. The dimension of that space is the count of rows in the column of numbers (arithmetic representation). Graphically, the dimension must be figured out in context, since all arrows look the same.
b. Scalar multiplication and addition in both representations.
c. Dot product
    i. takes two vectors, produces a *number*
    ii. arithmetical operation: componentwise multiply then add
    iii. interpreting graphically
        a. proportional to (cosine of) angle between vectors
d. Length of a vector
    i. arithmetically: pythagorus
    ii. arithmetically: sqruare root of dot product with itself
    iii. graphically: ruler.

```{r}
gvec(from=c(-2,3), to=c(4,4), color="red") %>%
graph_paper(xticks=-5:5, yticks=-5:5) %>%
    gvec(from=c(0,0), to=c(-2,3))
```
