---
title: "Exercises"
author: "Daniel Kaplan"
---

```{r include=FALSE, context="setup"}
library(etude2)
```

Short answer exercises are often arranged to give immediate feedback that goes beyond right-or-wrong. In other words, when an answer is incorrect, we try to tell you *why* it is incorrect to provide a hint to guide you to the correct answer. 

To see the feedback, place your cursor just after the end an answer choice. You'll notice that the cursor changes from an I to an arrow when you are in the right spot. For instance, in the following frivolous example , try placing your cursor just after the period in "The element calcium." It takes several seconds for the feedback to appear, so leave your cursor in place until you see the feedback.

```{r EI-1-1, echo=FALSE}
askMC(
  prompt = "What does the  word *calculus* mean literally?",
  "+A small stone+" = "Such stones were part of an apparatus for doing numerical calculations, back in the days before  the pencil was invented.",
  "A source of trouble to students." = "Calculus sometimes has that reputation, but as they say, \"No pain, no gain.\". Find a better answer.",
  "The element calcium." = "No, but when you get the right answer you'll realize the two are related in some cases."
)
```

Some exercises are meant as drill, and provide only right-or-wrong feedback. Still other exercises provide no feedback at all. Typically, these are intended to raise the stakes for graded work, encouraging you to think carefully before choosing an answer.

