---
id: "cow-likes-chair"
author: "Daniel Kaplan"
---

Since the inputs to functions are **quantities**, it's convenient to give them names that remind the human reader what they stand for.

The most common name we will use in this book for an input to a function will be:

- $\large x$ a general-purpose name that complies with the practice in high-school algebra. Remember, for us $x$ is not an "unknown" as it so often is in algebra; in calculus $x$ is the name of an input. 

Perhaps next most common as an input name is 

- $\large t$. This name is typically used when the input is meant to be ***time***. So if we were creating a function to represent the relationship between time (of day) and outdoor brightness, we might use this notation: $\text{brightness}(t)$

Other input names we will use in this book are $y$, $z$, $u$, $v$, $w$. But we won't shy away from more descriptive names, like $T$ for "temperature" or $V$ for volume, or even $\text{altitude}$ (which describes itself). 

For each of the following function definitions, what is the input name?

```{r clc2-1, echo=FALSE, results="markup"}
askMC(
  "Input name in $g(t) \\equiv 2 t^2 + 8$?",
  "+t+", "u", "v", "w", "x", "y", "z", random_answer_order=FALSE
)
```

```{r clc2-2, echo=FALSE, results="markup"}
askMC(
  "Input name in $\\line(z) \\equiv a z + b$?",
  "t", "u", "v", "w", "x", "y", "+z+", random_answer_order=FALSE
)
```

```{r clc2-3, echo=FALSE, results="markup"}
askMC(
  "Input name in $h(t) \\equiv 2 t^2 + 8 w$?",
  "+t+", "u", "v", "w", "x", "y", "z", random_answer_order=FALSE
)
```

```{r clc2-4, echo=FALSE, results="markup"}
askMC(
  "Input name in $f(u) \\equiv a u + b + u^2$?",
  "t", "+u+", "v", "w", "x", "y", "z", random_answer_order=FALSE
)
```

```{r clc2-5, echo=FALSE, results="markup"}
askMC(
  "Input name in $g(w) \\equiv x + 4$?",
  "t", "u", "v", "+w+", "x"="Tricked you! It's the name in the parentheses on the left side of $\\equiv$ that matters. Evidently, the output of $g(w)$ doesn't depend on $w$. That's not much of a relationship, but it is a common one.", "y", "z", random_answer_order=FALSE
)
```
