---
id: "beech-ride-bowl"
created: "Mon Jul 19 16:37:13 2021"
global_id: "Rdc7vE"
---

Every $n\times n$ matrix---we are working with $2\times 2$ matrices for simplicity--has associated with it a set of $n$ numbers called *eigenvalues*.  This is a mix of the German word "eigen" meaning "characteristic," and the English work "values." Sometime they are called *characteristic values* just to render them purely in English. But "eigen" has a single, specific meaning in mathematics, so the word "eigenvalue" is more evocative to mathematicians than is "characteristic values."

The letter used to represent eigenvalues is the Greek lambda: $\lambda$. Since our $2\times 2$ matrices have 2 eigenvalues, we often write the pair as $\lambda_1$ and $\lambda_2$ or even as $\lambda_{1,2}$. 

For an [ab10] matrix, the formula for the eigenvalues is

$$\lambda_{1,2} = \frac{1}{2} \left[a \pm \sqrt{a^2 + 4b\strut}\right]$$

Note the use of the $\pm$ in the formula. The same formula gives *both* eigenvalues, one from the $+$ branch of the $\pm$, the other from the $-$ branch. Using the $\pm$ rather than two different formulas---one with $-$ and one with $+$---is merely a convenience to make it explicit what the two formulas would have in common.

Writing $\lambda_{1,2}$ is nothing more than saying $\lambda_1$ *and* $\lambda_2$.

```{r char1-1, echo=FALSE, results="markup"}
etude2::etudeQ(
  "What are the numerical values of the eigenvalues of the matrix $\\left[\\begin{array}{rc}-1 & \\frac{3}{4} \\\\ 1 & 0\\end{array}\\right]$?",
  "+$\\frac{1}{2}$ and $-\\frac{3}{2}$+",
  "$\\frac{1}{2}$ and $-\\frac{5}{2}$",
  "$\\frac{1}{2}$ and $+\\frac{5}{2}$",
  "$-\\frac{1}{2}$ and $-\\frac{5}{2}$"
)
```

```{r char1-2, echo=FALSE, results="markup"}
etude2::etudeQ(
  "What are the numerical values of the eigenvalues of the matrix $\\left[\\begin{array}{rc}0 & \\frac{1}{2} \\\\ 1 & 0\\end{array}\\right]$?",
  "+$\\pm\\sqrt{2}/2$+",
  "$\\pm\\sqrt{2}$",
  "$\\pm 2$",
  "$\\pm\\frac{1}{2}$"
)
```

Eigenvalues are important because they dramatically simplify the interpretation of matrices and of dynamical system behavior near fixed points.

Eigenvalues are numbers. To use the language of linear combinations, an eigenvalue is a *scalar*. As you know, when you multiply a vector times a scalar, the result has the same direction (or the exact opposite direction) but might be longer or shorter. Like this:
$$ 3 \left[\begin{array}{c}1\\2 \end{array}\right] = \left[\begin{array}{c}3\\6 \end{array}\right]\ \ \ \ \text{or}\ \ \ \ \ 
-4 \left[\begin{array}{r}3\\-1 \end{array}\right] = \left[\begin{array}{r}-12\\4\end{array}\right]$$
In contrast, **matrix** multiplication of a vector can generate a result that has a *different direction* than the original vector, like this:

$$\left[\begin{array}{rc}0 & \frac{1}{2} \\ 1 & 0\end{array}\right]\cdot \left[\begin{array}{c}2 \\ 1\end{array}\right] =
\left[\begin{array}{c}\frac{1}{2} \\2\end{array}\right]\ \ \ \ \text{or}\ \ \ \ \left[\begin{array}{rc}-1 & \frac{3}{4} \\ 1 & 0\end{array}\right]\cdot \left[\begin{array}{r}1 \\ -4\end{array}\right] =
\left[\begin{array}{c}4 \\1\end{array}\right]$$

An **eigenvector** is a vector has a seemingly magical property with respect to multiplication by the matrix it characterizes: the result of the multiplication is a vector that has the same direction as the original but scaled by it's corresponding eigenvalue $\lambda$. That is, if $\lambda$ is an eigenvalue of ${\mathbf M}$ and ${\mathbf \Lambda}$ is the corresponding eigenvector: 
$${\mathbf M}\cdot {\mathbf \Lambda} = \lambda {\mathbf \Lambda}$$

```{r char1-3, echo=FALSE, results="markup"}
etude2::etudeQ(
  "Which of these is an eigenvector of the matrix $\\left[\\begin{array}{rc}-1 & \\frac{3}{4} \\\\ 1 & 0\\end{array}\\right]$? (Hint: Multiply the matrix by each of the vectors listed. For one of them, the result of the multiplication will be aligned with the vector used. You can use the constant of proportionality to answer the next question, but you'll have to start by choosing the right one of the options that follow.)",
  "+$\\left[\\begin{array}{r}-3\\\\2\\end{array}\\right]$+",
  "$\\left[\\begin{array}{r}3\\\\2\\end{array}\\right]$",
  "$\\left[\\begin{array}{r}-2\\\\3\\end{array}\\right]$", 
  "$\\left[\\begin{array}{r}2\\\\3\\end{array}\\right]$"
)
```

```{r char1-4, echo=FALSE, results="markup"}
etude2::etudeQ(
  "What is the eigenvalue associated with the correct answer from the previous problem?",
  "+-1.5+",
  "-1",
  "-0.5",
  "0.5", 
  "1.5",
  random_answer_order = TRUE
)
```

```{r char1-5, echo=FALSE, results="markup"}
etude2::etudeQ(
  "Consider the two matrices $${\\mathbf M}_1 = \\left[\\begin{array}{rr}5 & 6 \\\\0 & 1\\end{array}\\right]\\ \\ \\text{and}\\ \\ \\left[\\begin{array}{rr}-3 & 1 \\\\0 & 1\\end{array}\\right]$$ Which of the following is **not** an eigenvector of ${\\mathbf M}_1$ or of ${\\mathbf M}_2$? (We've rounded off to two decimal places and provided a sandbox for calculations.)",
  "$\\left[\\begin{array}{r}6\\\\1\\end{array}\\right]$",
  "$\\left[\\begin{array}{r}-1\\\\1\\end{array}\\right]$",
  "+$\\left[\\begin{array}{r}3\\\\2\\end{array}\\right]$+",
  "$\\left[\\begin{array}{r}3.02\\\\10\\end{array}\\right]$",
  "$\\left[\\begin{array}{r}3.30\\\\-1\\end{array}\\right]$"
)
```

You can use the sandbox below either for arithmetic calculations that you hand write or to use matrix-related R commands. In the initial code in the sandbox is a test that shows whether a vector `v` is an eigenvector of `M`. In this example it is and has eigenvalue 0.5615 (rounding off to 4 decimal places). You can print out `M` or `v` in the ordinary way, with the simple commands `M` or `v` respectively.
```{r char1-6, exercise=TRUE, exercise.cap="Matrix calculations", exercise.nlines=7, eval=FALSE}
M <- rbind(cbind(-3, 2), cbind(1, 0))
v <- rbind(5.6155, 10)
(M %*% v) / v # element-by-element division

```

