---
id: "walnut-tell-pot"
created: "Sun Dec  5 16:41:45 2021"
global_id: "iRZfBY"
---

# `r insert_calcZ_exercise("XX.XX", "iRZfBY", "Exercises/walnut-tell-pot.Rmd")`

Use Block 4 ape-choose-closet.Rmd this to explore the importance of $dt$ in an exercise for Block 6.

Here is the commentary that went with ape-choose-closet:

In the code originally in the sandbox, the trajectory was constructed with 100 Euler steps; the time interval runs from $t=0$ to $t=10$ and $dt = 0.1$. Practical optimization algorithms are designed to take steps that are as large as feasible. But if the steps are too large, the trajectory can overshoot the target. Dealing with this problem is a major feature of optimization algorithms and is non-trivial.

Purely to illustrate, we can investigate stepsize by changing `dt` in the two `integrateODE()` expressions. 

```{r fg1-4, eval=FALSE, echo=FALSE, results="markup"}
askMC(
  "What happens when you set `dt=1.0`?",
  "Both trajectories smoothly curve around and reach the maximum or minimum.",
  "+The trajectories reach the maximum or minimum, but in a clunky, jerkwise fashion.+",
  "The trajectories oscillate around the maximum or minimum without settling down.",
  "Everything falls apart and the trajectories leave the domain of interest.",
  random_answer_order = FALSE
)
```

```{r fg1-5, eval=FALSE, echo=FALSE, results="markup"}
askMC(
  "What happens when you set `dt=10.0`?",
  "Both trajectories smoothly curve around and reach the maximum or minimum.",
  "The trajectories reach the maximum or minimum, but in a clunky, jerkwise fashion.",
  "The trajectories oscillate around the maximum or minimum without settling down.",
  "+Everything falls apart and the trajectories leave the domain of interest.+",
  random_answer_order = FALSE
)
```


