---
id: "first-fundamental-theorem"
author: "Daniel Kaplan"
---


Remember our conventions for notation:

- Numerical constants (perhaps with units)
    - Symbols: e.g. $a$, $b$, $c$, $x_0$, $t^{\star}$
    - Examples: 3.2, 4.8 meters, 17 feet/sec$^2$
- Quantities that can vary (perhaps with units)
    - Symbols: e.g. $x$, $t$, $y$, $u$, $v$
    - Examples: position, time, velocity
- Functions of a variable
    - Symbols: e.g. $f(x)$, $g(t)$, $h(x, t)$
    - Examples: position as a function of time, density as a function of position
- Functions evaluated at a specific numerical input
    - Symbols: e.g. $f(a)$, $g(t_0)$, $h(x^{\star}, t^{\star})$
    - Examples: velocity at the finish line, starting position

In particular, take care to distinguish between these two kinds of symbolic items:

1. $f(x)$, $f()$ as a function of $x$
2. $f(x_0)$, the function $f()$ evaluated at the specific input $x_0$, producing (typically) a quantity like 3.5 meters/sec.

A major source of confusion for students is that $a$ is a constant, even though we are not yet saying specifically which numerical value that constant has. Think of $a$ as meaning "insert constant here." In terms of derivatives ...

LIST TAKEN OUT TO GET TUFTE TO WORK

<!--
i. $\partial_x f(a) = 0$
ii. $\partial_x f(x)$ is a function
iii. $\partial_u f(x) = 0$, since $u$ and $x$ are different variables.
iv. $\partial_u f(u)$ is a function, the exact same function as in (ii).

-->

With this in mind, turn to our three perspectives on a definite integral $$\int_a^b f(x) dx \ \ =\ \  \left. F(x) \right|_a^b\ \  =\ \  F(b) - F(a)$$

- $a$ and $b$ are numerical constants
-$f(x)$  and $F(x)$ are functions of $x$
-$F(a)$ is the function $F()$ evaluated at the specific input $a$, producing a quantity. Likewise $F(b)$.

```{r fft1, echo=FALSE}
askMC(
  "What kind of a thing is $F(u)$, according to our notation convention?",
"a number" = "Is $u$ a number in our notation?",
"a function of $x$" = "Close, but not $x$.",
"+a function of $u$+",
"a definite integral" = "A definite integral would be something like $\\int_a^b f(x) dx$ or $\\left. F(x) \\right|_a^b$ or $F(b) - F(a)$."
  )
```

```{r fft2, echo=FALSE}
explain <- "A function always has one or more variables as input. $a$ is not a variable but a numerical constant."
askMC(
  "What kind of a thing is $F(a)$, according to our notation convention?",
"+a number+" = "We're evaluating $F()$ at a specific numerical value of the input, namely $a$.",
"a function of $x$" = explain,
"a function of $u$" = explain,
"a definite integral" = "A definite integral would be something like $\\int_a^b f(x) dx$ or $\\left. F(x) \\right|_a^b$ or $F(b) - F(a)$."
  )
```

```{r fft3, echo=FALSE}
askMC(
  "What kind of a thing is $F(u) - F(a)$, according to our notation convention?",
"a number" = "Is $u$ a number in our notation?",
"a function of $x$" = "Close, but not $x$.",
"+a function of $u$+" = "Since $F(u)$ is a function of $u$, and $F(a)$ is a constant, $F(u) - F(a)$ is a function of $u$ minus a constant, or, in other words, a function of $u$.",
"a definite integral" = "A definite integral would be something like $\\int_a^b f(x) dx$ or $\\left. F(x) \\right|_a^b$ or $F(b) - F(a)$."
  )
```


```{r fft4, echo=FALSE}
askMC(
  "According to our notation convention, what kind of a thing is $$\\int_a^u f(x) dx \\text{?}$$, ",
"a number",
"a function of both x and u",
"a function of $x$" = "Close, but not $x$.",
"+a function of $u$+" = "You can see this by referring to the three perspectives equation. $$\\int_a^u f(x)dx = F(u) - F(a)$$ which is a function of $u$.",
"a definite integral" = "Certainly it looks a lot like a definite integral. But $u$ is a variable, not a constant."
  )
```

```{r fft5, echo=FALSE}
askMC(
  "According to our notation convention, what kind of a thing is $$\\int_u^b f(x) dx \\text{?}$$, ",
"a number",
"a function of both x and u",
"a function of $x$" = "Close, but not $x$.",
"+a function of $u$+" = "You can see this by referring to the three perspectives equation. $$\\int_u^b f(x)dx = F(b) - F(u)$$ which is a function of $u$.",
"a definite integral" = "Certainly it looks a lot like a definite integral. But $u$ is a variable, not a constant."
  )
```


```{r fft6, echo=FALSE}
askMC(
  "According to our notation convention, what kind of a thing is $$\\int_u^x f(x) dx \\text{?}$$, ",
"a number",
"+a function of both x and u+" = "You can see this by referring to the three perspectives equation. $$\\int_u^x f(x)dx = F(x) - F(u)$$ which is a function of both $x$ and $u$.",
"a function of $x$",
"a function of $u$",
"a definite integral" = "Certainly it looks a lot like a definite integral. But $x$ and  $u$ are variables, not constants."
  )
```

Now turn to the entities involved in the so-called "First Fundamental Theorem of Calculus." ("Fundamental theorem" is a highfalutin way of saying something like, "This isn't obvious at first glance, and so you should be especially careful to memorize it so that you identify it when you see it." Another way to state it is, "Every function is the derivative of some anti-derivative." But you knew that already, since "every function has an anti-derivative.")

Here are the entities involved, which you'll recognize as a slight modification of an earlier statement:

$$\partial_u \int_a^u f(x)dx \ \ =\ \ \partial_u \left. F(x) \right|_a^u \ \ = \ \ \partial_u \left(F(u) - F(a)\right) .$$
Let's look at the right-most expression $\partial_u \left(F(u) + F(a)\right)$ and exploit the the derivative of a sum is the sum of the derivatives. So ...
$$\partial_u \left(F(u) + F(a)\right) = \partial_u F(u) - \partial_u F(a) = \partial_u F(u)$$

```{r fft7, echo=FALSE}
askMC(
  "Which of the following correctly justifies the step $$\\partial_u F(u) - \\partial_u F(a) = \\partial_u F(u)\\ \\text{?}$$",
  "+$F(a)$ is a constant+" = "And the derivative of a constant is always 0.",
  "$F()$ is an anti-derivative.",
  "$F(b)$ does not appear.",
  "$F(u) = \\int f(x) dx$"
)
```

Taking the left-most and right-most expressions in the above equation, we have
$$\partial_u \int_a^u f(x) dx = \partial_u F(u)$$

```{r fft8, echo=FALSE}
askMC(
  "Is there an algebraic simplification of $\\partial_u F(u)$?",
  "No, because it depends on what $F(u)$ is." = "But we know what $F(u)$ is: the anti-derivative of $f(u)$.",
  "+Yes, because $\\partial_u F(u)$ is simply $f(u)$.+",
  "No, because we could just as easily have written $\\partial_x F(x)$",
  "Yes, because it is the same thing as $\\partial_x F(x)$"
)
```

The equation $$\partial_u \int_a^u f(x) dx \ \ = \ \  f(u)$$ means that "differentiation undoes integration" or, as we've been putting it, "differentiation undoes anti-differentiation."
