---
id: "horse-takes-pillow"
author: "Daniel Kaplan"
global-id: "NYPDR"
---

```{r echo=FALSE}
f1 <- rfun(~ x, seed=782)
df1 <- D(f1(x) ~ x)
slice_plot(df1(x) ~ x, domain(x=c(-5,5)))
```

The graph shows a function $f(x)$. Like all functions, it has an anti-derivative. Here are several functions, $F_1(x)$ through $F_4(x)$. One and only one is the anti-derivative of the function shown in the graph. Your job is to figure out which one.

```{r}
F1 <- rfun(~ x, seed = 782)
F2 <- rfun(~ x, seed = 934)
F3 <- rfun(~ x, seed = 297)
F4 <- rfun(~ x, seed = 892)
```

Don't worry about how `rfun()` makes a function. Once you run those four lines, you can use `slice_plot()` in the usual way to plot out any of the four functions over the domain $-5 \leq x \leq 5$. 

```{r horse-takes-pillow-1, echo=FALSE, results="markup"}
askMC(
  "Which one of the functions is an anti-derivative of the function graphed above? (Hint: See the next problem.)",
  "+`F1()`+", "`F2()`", "`F3()`", "`F4()`",
  random_answer_order = FALSE
)
```

```{r horse-takes-pillow-2, echo=FALSE, results="markup"}
askMC(
  "Confirming graphically that a candidate $F(x)$ is indeed the anti-derivative of $f(x)$ is hard because the two functions can have very different shapes. But there is a very simple operation you can apply to $F(x)$ to make the comparison trivial. What is this operation?",
  "anti-differentiate $F(x)",
  "+differentiate $F(x)$+" = "If the result matches the shape of $f(x)$, you've confirmed that $F(x)$ is the anti-derivative of $f(x)$.",
  "compose $f(x)$ with itself",
  "multiply $F(x)$ times $f(x)"
)
```

